### 2017-09-29 pkg/model4you/inst/JORS/man_model4you.Rnw
comments Torsten
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1596 255f4249-2263-4417-b1ea-8498779e27a7

%\textcolor{blue}{An overview of the software, how it was produced, and the research for which it has been used, including references to relevant research articles. A short comparison with software which implements similar functionality should be included in this section. }
Studies in various fields randomly assign individuals to one of two groups with
different exposure and then measure a response. For example, in clinical trials
patients are assigned to one of two treatment groups where usually one
treatment group receives a new treatment or drug and the other treatment group
receives the standard of care or a placebo.

@@

\paragraph{(1) Model-based trees} identify subgroups where within the subgroups
the model parameters are similar and between groups the model parameters are
different. This is achieved by finding instabilities in the model parameters
with respect to a variable (characteristicum) and recursively partitioning the
data into groups. If, for example the algorithm finds that men and women have
differing treatment effects, the data is partitioned into two subgroups.
Details on model-based trees in general can be found in
\citep{zeileis_model-based_2008} and for the special use case for stratified
treatment effect estimation in \citep{seibold_model-based_2016}.  Just a single
line of code lets the user compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@

\paragraph{(2) Personalised models} use model-based random forests to estimate
similarity of individuals in terms of model parameters. For each individual a
personalised model can be estimated based on a weighted set of the original
data, where the similarity measure corresponds to the weight. Details on the
personalised models can be found in \citep{seibold_individual_2017}.  
Computing personalised models for all observations in the training data is
simple:
<<eval=FALSE>>=
pm_forest <- pmforest(base_model)
pers_models <- pmodel(pm_forest)
@

@@

\paragraph{Mathematics exam analysis:}
In 2014 first-year business and economics students at the University of Innsbruck
were divided into two examination groups. Group 1 wrote the exam in the morning and
group 2 started after the first group finished. The exams for the two groups were 
slightly different. Translated into clinical trial language, the exam groups are
the treatment groups and the two different exams are the treatments.
The data can be accessed and prepared as follows:

@@

The tree (see Figure~\ref{fig:math_tree}) divides students based on the
percentage of successful online \Sexpr{'tests'}. These online tests were
conducted biweekly thrughout the semester.  The largest difference between the
two exam groups is in the students who did very well in the online tests (more
than \Sexpr{round(tr_math[[5]]$node$split$breaks, 1)} percent correct). The
tree thus gives us much more information on the fairness of the exam than the
simple linear model, which is that it does not seem to be fair for students who
did very well thrughout the semester (at this point we should state that the
students self selected into the two exam groups which might also be the reason
for differences in exam performance).

@@

\section*{Implementation and architecture}
\begin{figure}
\centering
\include{treefamily}
\caption{The family of model-based trees and forests. Square nodes represent
functions, round nodes represent function arguments. Solid lines represent
direct usage, dashed lines indirect usage (iteration between tree and model
estimation), and dotted lines represent close similarity but no dependency.}
\label{fig:family}
\end{figure}

The R package \emph{model4you} is based on the \emph{partykit} package. In
fact, everything that can be done with \emph{model4you} can already be done
with \emph{partykit} (version $\geq$ 1.2-0). \emph{model4you} is merely a
convenience interface and focusses on ease of use and interpretability. Various
other packages depend on \emph{partykit}. Figure~\ref{fig:family} shows all
packages depending on \emph{partykit} that deal with model-based trees and
forests. Functions \Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::mob()'}
provide the basis for all others. Note that \Sexpr{'partykit::ctree()'}
computes conditional inference trees but by specifying argument
\Sexpr{'ytrafo'} appropriately, it can also compute model-based trees. This
functionality is used for packages \emph{disttree} (partitioning of Generalized
Additive Models for Location, Scale and Shape), \emph{trtf} (partitioning of
transformation models) and \emph{model4you}.  Many of the packages shown in
Figure~\ref{fig:family} are available on CRAN.  The remaining packages are
under development and available on
\url{https://r-forge.r-project.org/projects/partykit/}.

>>>>

%\textcolor{blue}{An overview of the software, how it was produced, and the research for which it has been used, including references to relevant research articles. A short comparison with software which implements similar functionality should be included in this section. }
Studies in various fields randomly assign individuals to one of two groups with
different exposure and then measure a response. For example, in clinical trials
patients are assigned to one of two treatment groups where usually one
treatment group receives a new treatment or drug and the other treatment group
receives the standard of care or a placebo. Other examples are in A-B testing
in marketing studies or any other two group comparisons such as the mathematics
exam discussed below, where students were divided into different exam groups
and received slightly different exam tasks. In the following we will refer to
the two groups as \emph{treatment groups} and to the group indicator as
\emph{treatment indicator}, which always takes values $0$
(individual in first group) and $1$ (individual in second group).

@@

\paragraph{(1) Model-based trees} identify subgroups where within the subgroups
the model parameters are similar and between groups the model parameters are
different. This is achieved by finding instabilities in the model parameters
with respect to a variable (characteristicum) and recursively partitioning the
data into groups. If, for example the algorithm finds that men and women have
differing treatment effects, the data is partitioned into two subgroups.
Details on model-based trees in general can be found in
\citep{zeileis_model-based_2008} and for the special use case for stratified
treatment effect estimation in \citep{seibold_model-based_2016}.  Just a single
line of code lets the user compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@
Note that \Sexpr{'pmtree()'} uses the data given in the call of the base 
model. It automatically uses variables not used in the model formula 
(in the example above \Sexpr{'response ~ treatment'}) as potential subgroup 
defining variables. This can be edited using the \Sexpr{'zformula'} argument.

\paragraph{(2) Personalised models} use model-based random forests to estimate
similarity of individuals in terms of model parameters. For each individual a
personalised model can be estimated based on a weighted set of the original
data, where the similarity measure corresponds to the weight. Details on the
personalised models can be found in \citep{seibold_individual_2017}.  
Computing personalised models for all observations in the training data is
simple:
<<eval=FALSE>>=
pm_forest <- pmforest(base_model)
pers_models <- pmodel(pm_forest)
@ 
Again here the potential effect-modifying variables are taken by default
as all variable not given in the model formula and can be defined using 
the \Sexpr{'zformula'} argument in \Sexpr{'pmforest()'}.

@@

\paragraph{Mathematics exam analysis:}
In 2014 first-year business and economics students at the University of
Innsbruck were divided into two examination groups. Group 1 wrote the exam in
the morning and group 2 started after the first group finished. The exams for
the two groups were slightly different.  The data can be accessed and prepared
as follows:

@@

The tree (see Figure~\ref{fig:math_tree}) divides students based on the
percentage of successful online \Sexpr{'tests'}. These online tests were
conducted biweekly throughout the semester.  The largest difference between the
two exam groups is in the students who did very well in the online tests (more
than \Sexpr{round(tr_math[[5]]$node$split$breaks, 1)} percent correct). The
tree thus gives us much more information on the fairness of the exam than the
simple linear model, which is that it does not seem to be fair for students who
did very well thrughout the semester (at this point we should state that the
students self selected into the two exam groups which might also be the reason
for differences in exam performance).

@@

\section*{Implementation and architecture}
The R package \emph{model4you} is focussed on ease of use and interpretability.
Users can take a simple model that they know and understand as basis and simply
plug it into \Sexpr{'pmtree()'} or \Sexpr{'pmforest()'} depending on whether
they want subgroup wise or personalised models. The basis for these
functionalities is provided by the \emph{partykit} package which is a widely
used R package for trees and forests. The \emph{model4you} package provides
wrappers for the well implemented and tested functions
\Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::cforest()'} and extends the
functionalities where needed to improve usability and interpretability.

The \emph{partykit} package provides the basis for functionalities in other
packages namely \emph{glmertree}, \emph{psychotree}, \emph{betareg} (all on
CRAN), \emph{trtf}, \emph{disttree}, \emph{lagsarlmtree} and \emph{palmtree}
(all available on R-Forge with a plan to move them to CRAN).
